---
---

## Raster upon Raster

The [raster](){:.rlib} library uniformly handles two-dimensional data as a
`RasterLayer` object, created with the `raster()` function. There are several
cases where raster data has a third dimension.

- a multi-band image (e.g. Landsat scenes)
- closely related images (e.g. time series of NDVI values)
- three dimensional data (e.g. GCM products)

===

Any `RasterLayer` objects that have the same extent and resolution (i.e. cover
the same space with the same number of rows and columns) can be combined into
one of two container types:

- a `RasterStack` created with `stack()`
- a `RasterBrick` created with `brick()`

===

The main difference is that a `RasterStack` is loose collection of `RasterLayer` objects that can refer to different files (but must all have the same extent and resolution), whereas a `RasterBrick` can only point to a single file.

===

## Raster Stacks

The layers of a "stack" can refer to data from separate files, or even a mix of
data on disk and data in memory.

Read raster data and create a stack. 

```{r, handout = 0}
ndvi_yrly <- Sys.glob('data/r_ndvi_*.tif')
```
```{r}
ndvi_yrly
```
We read all the `.tif` files in the `data` folder using the `*` wildcard character.
In this case, we read the two `.tif` files that are in the `data` folder.
{:.notes}

===

```{r, handout = 0}
ndvi <- stack(ndvi_yrly)
names(ndvi) <- c(
  'Avg NDVI 2002',
  'Avg NDVI 2009')
```
Using the `stack()` function we create a raster stack and assign it to the `ndvi` object.
We name each layer in the stack by using the `names` function. These will also be the titles for the plots.
{:.notes}

```{r}
plot(ndvi)
```

===

The data source for the first layer is the first file in the list. Note that the
CRS is missing: it is quite possible to stack files with different
projections. They only have to share a common extent and resolution.

```{r}
# display metadata for the 1st raster in the ndvi stack
raster(ndvi, 1)
```

===

Set the CRS using the [EPSG code
3338](http://spatialreference.org/ref/epsg/nad83-alaska-albers/) for an Albers
Equal Area projection of Alaska using the NAD38 datum.

```{r, handout = 0}
crs(ndvi) <- '+init=epsg:3338'
```
```{r}
# display metadata for the ndvi stack
raster(ndvi, 0)
```

===

The `ndvi` object only takes up a tiny amount of memory. The pixel values
take up much more space on disk, as you can see in the file browser.

```{r}
print(object.size(ndvi),
  units = 'KB',
  standard = 'SI')
```

===

Why so small? The `ndvi` object is only metadata and a pointer to where the
pixel values are saved on disk.

```{r}
inMemory(ndvi)
```

Whereas `read.csv()` would load the named file into memory, the
[raster](){:.rlib} library handles files like a database where possible. The
values can be accessed, to make those plots for example, but are not held in
memory. This is the key to working with deep stacks of large rasters.
{:.notes}

===

## Wildfires in Alaska

Many and massive wildfires burned in Alaska and the Yukon between 2001 and 2009.
Three large fires that burned during this period (their locations are in a
shapefile) occurred within boreal forest areas of central Alaska.

```{r, handout = 0}
scar <- st_read(
  'data/OVERLAY_ID_83_399_144_TEST_BURNT_83_144_399_reclassed',
  crs = 3338)
plot(ndvi[[1]])
plot(st_geometry(scar), add = TRUE)
```
We read the `OVERLAY_ID_83_399_144_TEST_BURNT_83_144_399_reclassed` directory. This
is a polygon shapefile containing the geometry of a wildfire in central Alaska.
We assign the `EPSG:3338` CRS for Alaska Albers when reading the shapefile.
We plot the first raster layer in the `ndvi` stack and then plot an overlay of the
wildfire using the polygon shapefile. 
{:.notes}

===

For faster processing in this lesson, crop the NDVI imagery to the smaller
extent of the shapefile.

```{r, handout = 0}
burn_bbox <- st_bbox(scar)
ndvi <- crop(ndvi, burn_bbox)
```
Using `st_bbox()` we find the bounding box of the wildfire scar polygons. 
We assign it to `burn_bbox` and crop the raster stack object `ndvi` to that extent.
{:.notes}

```{r}
plot(ndvi[[1]], ext = burn_bbox)
plot(st_geometry(scar), add = TRUE)
```

===

Notice, however, that the NDVI values are now stored in memory. That's okay for
this part of the lesson; we'll see the alternative shortly.

```{r}
inMemory(ndvi)
```

===

## Pixel Change

Use element-wise subtraction to give a difference raster, where negative values
indicate a higher NDVI in 2002, or a decrease in NDVI from 2002 to 2009.

```{r, handout = 0}
diff_ndvi <- ndvi[[2]] - ndvi[[1]]
names(diff_ndvi) <- 'Difference'
```
`diff_ndvi` is a raster object containing the difference between the NDVI values for 2002 and 2009 for each pixel.
{:.notes}

```{r}
plot(diff_ndvi)
```

===

The histogram shows clearly that change in NDVI within this corner of Alaska clusters around two modes.

```{r}
hist(diff_ndvi)
```

===

One way to "classify" pixels as potentially affected by wildfire is to threshold
the difference. Pixels below `-0.1` mostly belong to the smaller mode, and may
represent impacts of wildfire.

```{r}
plot(diff_ndvi < -0.1)
plot(st_geometry(scar), add = TRUE)
```

===

The threshold value could also be defined in terms of the variability among
pixels.

Centering and scaling the pixel values requires computation of their
mean and standard variation. The `cellStats` function efficiently applies a few
common functions across large rasters, regardless of whether the values are in
memory or on disk.
{:.notes}

```{r, handout = 0}
diff_ndvi_mean <-
  cellStats(diff_ndvi, 'mean')
diff_ndvi_sd <-
  cellStats(diff_ndvi, 'sd')
```

===

Mathematical operations with rasters and scalars work as expected; scalar
values are repeated for each cell in the array.
{:.notes}

The difference threshold of `-0.1` appears roughly equivalent to a threshold of
1 standard deviation below zero.

In the following code block we standardize the NDVI difference by subtracting the
mean we calculated earlier, and then dividing by the standard deviation.
{:.notes}

```{r, handout = 0}
diff_ndvi_stdz <-
  (diff_ndvi - diff_ndvi_mean) /
  diff_ndvi_sd
names(diff_ndvi_stdz) <- 'Std. Diff.'
```
```{r}
hist(diff_ndvi_stdz, breaks = 20)
```

===

Standardizing the pixel values does not change the overall result.

```{r}
plot(diff_ndvi_stdz < -1)
plot(st_geometry(scar), add = TRUE)
```
